Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-02-20 paper/main.tex
add non-breaking spaces between 'lecture'/'quiz' and number so line doesn't start with digit

We asked participants in our study to complete brief multiple-choice quizzes
before, between, and after watching two lecture videos from the Khan
Academy~\citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture
video, entitled \textit{Four Fundamental Forces}, discussed the four
fundamental forces in physics: gravity, strong and weak interactions, and
electromagnetism. The second, entitled \textit{Birth of Stars}, provided an
overview of our current understanding of how stars form. We selected these
particular lectures to satisfy three general criteria. First, we wanted both
lectures to be accessible to a broad audience (i.e., with minimal prerequisite
knowledge) so as to limit the impact of prior training on our participants'
abilities to learn from the lectures. To this end, we selected two introductory
videos that were intended to be viewed at the start of students' training in
their respective content areas. Second, we wanted both lectures to have some
related content, so that we could test our approach's ability to distinguish
similar conceptual content. To this end, we chose two videos from the same (per
instructor annotations) Khan Academy course domain, ``Cosmology and
Astronomy.'' Third, we sought to minimize dependencies and specific overlap
between the videos. For example, we did not want participants' abilities to
understand one video to (directly) influence their abilities to understand the
other. To satisfy this last criterion, we chose videos from two different
lecture series (lectures 1 and 2 were from the ``Scale of the Universe'' and
``Stars, Black Holes, and Galaxies'' series, respectively).

@@

Of course, even though the timecourses in
Figure~\ref{fig:knowledge-timeseries}A and C provide detailed
\textit{estimates} about participants' knowlege, those estimates are only
\textit{useful} to the extent that they accurately reflect what participants
actually know. As one sanity check, we anticipated that the knowledge estimates
should show a content-specific ``boost'' in participants' knowledge after
watching each lecture. In other words, if participants learn about each
lecture's content when they watch each lecture, the knowledge estimates should
reflect that. After watching the \textit{Four Fundamental Forces} lecture,
participants should show more knowledge for the content of that lecture than
they had before, and that knowledge should persist for the remainder of the
experiment. Specifically, knowledge about that lecture's content should be
relatively low when estimated using Quiz~1 responses, but should increase when
estimated using Quiz~2 or~3 responses (Fig.~\ref{fig:knowledge-timeseries}B).
Indeed, we found that participants' estimated knowledge about the content of
the \textit{Four Fundamental Forces} was substantially higher on Quiz~2 versus
Quiz~1 ($t(49) = 8.764,~p < 0.001$) and on Quiz~3 versus Quiz~1 ($t(49) =
10.519,~p < 0.001$). We found no reliable differences in estimated knowledge
about that lecture's content on Quiz~2 versus~3 ($t(49) = 0.160,~p = 0.874$).
Similarly, we hypothesized (and subsequently confirmed) that participants
should show more estimated knowledge about the content of the \textit{Birth of
Stars} lecture after (versus before) watching it
(Fig.~\ref{fig:knowledge-timeseries}D). Specifically, since participants
watched that lecture after taking Quiz~2 (but before Quiz~3), we hypothesized
that their knowledge estimates should be relatively low on Quizzes 1 and 2, but
should show a ``boost'' on Quiz~3. Consistent with this prediction, we found no
reliable differences in estimated knowledge about the \textit{Birth of Stars}
lecture content on Quizzes 1 versus 2 ($t(49) = 1.013,~p = 0.316$), but the
estimated knowledge was substantially higher on Quiz~3 versus~2 ($t(49) =
10.561,~p < 0.001$) and Quiz~3 versus 1 ($t(49) = 8.969,~p < 0.001$).

@@

    \caption{\textbf{Mapping out the geometry of knowledge and learning.}
    \textbf{A. Average ``knowledge maps'' estimated using each quiz.} Each map
    displays a 2D projection of the estimated knowledge about the content
    reflected by \textit{all} regions of topic space (see
    \nameref{subsec:knowledge-maps}). The topic trajectories of each
    lecture are indicated by dotted lines, and the coordinates of each question are indicated by
    dots. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,
    and~\individualKnowledgeMapsC. \textbf{B. Average ``learning maps''
    estimated between each successive pair of quizzes.} The learning maps follow
    the same general format as the knowledge maps in Panel A, but here the shading at each
    coordinate indicates the \textit{difference} between
    the corresponding coordinates in the indicated \textit{pair} of knowledge
    maps---i.e., how much the estimated knowledge ``changed'' between the two
    quizzes. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualLearningMapsA~and~\individualLearningMapsB. \textbf{C.
    Word clouds for sampled points in topic space.} Each word cloud displays
    the relative weights of each word (via their relative sizes) reflected by the blend of topics
    represented at the locations of the stars on the maps. The words' colors
    indicate how much each word is weighted, on average, across all timepoints'
    topic vectors in the \textit{Four Fundamental Forces} (bluer coloring) and
    \textit{Birth of Stars} (greener coloring) videos, respectively.}

@@

Over the course of the experiment, participants completed three 13-question
multiple-choice quizzes: the first before viewing lecture~1, the second between
lectures 1 and 2, and the third after viewing lecture~2
(Fig.~\ref{fig:experiment}). The questions appearing on each quiz, for each
participant, were randomly chosen from the full set of 39, with the constraints
that (a) each quiz contain 5 questions about lecture~1, 5 questions about
lecture~2, and 3 questions about general physics knowledge, and (b) each
question appear exactly once for each participant. The orders of questions on
each quiz, and the orders of answer options for each question, were also
randomized. Our experimental protocol was approved by the Committee for the
Protection of Human Subjects at Dartmouth College. We used the experiment to
develop and test our computational framework for estimating knowledge and
learning.

@@

After fitting a topic model to the two videos' transcripts, we could use the
trained model to transform arbitrary (potentially new) documents into
$k$-dimensional topic vectors. A convenient property of these topic vectors is
that documents that reflect similar blends of topics (i.e., documents that
reflect similar themes, according to the model) will yield similar coordinates
(in terms of correlation, cosine similarity, Kullback-Leibler divergence,
Euclidean distance, or other geometric measures). In general, the similarity
between different documents' topic vectors may be used to characterize the
similarity in conceptual content between the documents.

>>>>

We asked participants in our study to complete brief multiple-choice quizzes
before, between, and after watching two lecture videos from the Khan
Academy~\citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture
video, entitled \textit{Four Fundamental Forces}, discussed the four
fundamental forces in physics: gravity, strong and weak interactions, and
electromagnetism. The second, entitled \textit{Birth of Stars}, provided an
overview of our current understanding of how stars form. We selected these
particular lectures to satisfy three general criteria. First, we wanted both
lectures to be accessible to a broad audience (i.e., with minimal prerequisite
knowledge) so as to limit the impact of prior training on our participants'
abilities to learn from the lectures. To this end, we selected two introductory
videos that were intended to be viewed at the start of students' training in
their respective content areas. Second, we wanted both lectures to have some
related content, so that we could test our approach's ability to distinguish
similar conceptual content. To this end, we chose two videos from the same (per
instructor annotations) Khan Academy course domain, ``Cosmology and
Astronomy.'' Third, we sought to minimize dependencies and specific overlap
between the videos. For example, we did not want participants' abilities to
understand one video to (directly) influence their abilities to understand the
other. To satisfy this last criterion, we chose videos from two different
lecture series (lectures~1 and 2 were from the ``Scale of the Universe'' and
``Stars, Black Holes, and Galaxies'' series, respectively).

@@

Of course, even though the timecourses in
Figure~\ref{fig:knowledge-timeseries}A and C provide detailed
\textit{estimates} about participants' knowlege, those estimates are only
\textit{useful} to the extent that they accurately reflect what participants
actually know. As one sanity check, we anticipated that the knowledge estimates
should show a content-specific ``boost'' in participants' knowledge after
watching each lecture. In other words, if participants learn about each
lecture's content when they watch each lecture, the knowledge estimates should
reflect that. After watching the \textit{Four Fundamental Forces} lecture,
participants should show more knowledge for the content of that lecture than
they had before, and that knowledge should persist for the remainder of the
experiment. Specifically, knowledge about that lecture's content should be
relatively low when estimated using Quiz~1 responses, but should increase when
estimated using Quiz~2 or 3 responses (Fig.~\ref{fig:knowledge-timeseries}B).
Indeed, we found that participants' estimated knowledge about the content of
the \textit{Four Fundamental Forces} was substantially higher on Quiz~2 versus
Quiz~1 ($t(49) = 8.764,~p < 0.001$) and on Quiz~3 versus Quiz~1 ($t(49) =
10.519,~p < 0.001$). We found no reliable differences in estimated knowledge
about that lecture's content on Quiz~2 versus 3 ($t(49) = 0.160,~p = 0.874$).
Similarly, we hypothesized (and subsequently confirmed) that participants
should show more estimated knowledge about the content of the \textit{Birth of
Stars} lecture after (versus before) watching it
(Fig.~\ref{fig:knowledge-timeseries}D). Specifically, since participants
watched that lecture after taking Quiz~2 (but before Quiz~3), we hypothesized
that their knowledge estimates should be relatively low on Quizzes~1 and 2, but
should show a ``boost'' on Quiz~3. Consistent with this prediction, we found no
reliable differences in estimated knowledge about the \textit{Birth of Stars}
lecture content on Quizzes~1 versus 2 ($t(49) = 1.013,~p = 0.316$), but the
estimated knowledge was substantially higher on Quiz~3 versus 2 ($t(49) =
10.561,~p < 0.001$) and Quiz~3 versus 1 ($t(49) = 8.969,~p < 0.001$).

@@

    \caption{\textbf{Mapping out the geometry of knowledge and learning.}
    \textbf{A. Average ``knowledge maps'' estimated using each quiz.} Each map
    displays a 2D projection of the estimated knowledge about the content
    reflected by \textit{all} regions of topic space (see
    \nameref{subsec:knowledge-maps}). The topic trajectories of each
    lecture are indicated by dotted lines, and the coordinates of each question are indicated by
    dots. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,
    and~\individualKnowledgeMapsC. \textbf{B. Average ``learning maps''
    estimated between each successive pair of quizzes.} The learning maps follow
    the same general format as the knowledge maps in Panel A, but here the shading at each
    coordinate indicates the \textit{difference} between
    the corresponding coordinates in the indicated \textit{pair} of knowledge
    maps---i.e., how much the estimated knowledge ``changed'' between the two
    quizzes. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualLearningMapsA~and~\individualLearningMapsB. \textbf{C.
    Word clouds for sampled points in topic space.} Each word cloud displays
    the relative weights of each word (via their relative sizes) reflected by the blend of topics
    represented at the locations of the stars on the maps. The words' colors
    indicate how much each word is weighted, on average, across all timepoints'
    topic vectors in the \textit{Four Fundamental Forces} (blue) and
    \textit{Birth of Stars} (green) videos, respectively.}

@@

Over the course of the experiment, participants completed three 13-question
multiple-choice quizzes: the first before viewing lecture~1, the second between
lectures~1 and 2, and the third after viewing lecture~2
(Fig.~\ref{fig:experiment}). The questions appearing on each quiz, for each
participant, were randomly chosen from the full set of 39, with the constraints
that (a) each quiz contain 5 questions about lecture~1, 5 questions about
lecture~2, and 3 questions about general physics knowledge, and (b) each
question appear exactly once for each participant. The orders of questions on
each quiz, and the orders of answer options for each question, were also
randomized. Our experimental protocol was approved by the Committee for the
Protection of Human Subjects at Dartmouth College. We used the experiment to
develop and test our computational framework for estimating knowledge and
learning.

@@

After fitting a topic model to the two videos' transcripts, we could use the
trained model to transform arbitrary (potentially new) documents into
$k$-dimensional topic vectors. A convenient property of these topic vectors is
that documents that reflect similar blends of topics (i.e., documents that
reflect similar themes, according to the model) will yield similar coordinates
(in terms of Euclidean distance, correlation, or other geometric measures). In
general, the similarity between different documents' topic vectors may be used
to characterize the similarity in conceptual content between the documents.
